---
title: "20200828_Pathak_Vinayak_HW1"
author: "Vinayak Pathak2"
date: "8/28/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r}
df <- read.csv2(file ="Chr20ex.csv", sep = "\t", header = FALSE)
```

```{r}
summary(df)
```

```{r}
head(df)
```

```{r}
dim(df)
```

So there are 500 rows in the dataframe, i.e. 500 records
```{r}
df_pass <- df %>% filter(V7 == "PASS")
```

```{r}
head(df_pass)
```

```{r}
count(df_pass)[1, ]
```
Therefore total 500 records within the PASS Variable
```{r}
df_c_t <- df %>% 
  filter(V4 == "C", V5 =='T')
```

```{r}
count(df_c_t)[1, ]
```
There for total 116 records withh C->T SNPs in the Data
```{r}
head(df_c_t)
```
Extracting the correct rows as tibble and then plotting the graph
Renaming the columns


Plotting the chromossome histogram
```{r}
g <- ggplot(df, aes(class(V5)))

```


```{r}
g+geom_histogram()
```


```{bash eval=FALSE}
wget https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data
```
```{r}
df_hd <-read.table(file = "processed.cleveland.data", header =TRUE, sep =",")
```


```{r}
list_attrib <- c("age", "sex", "cp", "trestbps", "chol", "fbs", "restecg", "thalach", "exang", "oldpeak", "slope", "ca", "thal", "num")
```

```{r}
col_df = colnames(df_hd)
```

Renaming the columns of the dataframe

```{r}
#for(i in 2:length(col_df))

names(df_hd) <- list_attrib

```

```{r}
col_df_mod <-colnames(df_hd)
col_df_mod
```

```{r}

head(df_hd)
```

```{r}
dim(df_hd)
```

Total number of rows and columns in the the data frame are 302 and 14; therefore there are toatal 302 records and 14 attributes for eac hof the record.
```{r}
df_hd <- mutate(df_hd, diagnosed = (num>0))
```
An example of the display...
```{r}
head(df_hd) %>%
  select(age, sex, diagnosed)
```
```{r eval = FALSE}
```

Testing a general column
```{r eval}
summary(df_hd %>%
          select(age))
```
```{r}
temp <- df_hd %>%
  select(age)
```

```{r eval=FALSE}
summarize_column <- function(df, column_name) {
  if (!column_name %in% names(df)) {
    stop("Please enter the correct column name")
  }


}
```
```{r}
df_temp <- df_hd %>%
          select(age)
```
```{r}
(df_temp)
```
Following is the function to be defined...
```{r eval=FALSE}
summarize_column <- function(dfn, column_name) {
  if (!column_name %in% names(dfn)) {
    stop(cat("The queried column name", column_name, "is not a valid dataframe attribute please enter the correct column name"))
  }
temp <- summary(dfn 
        %>%select(column_name))

print(temp)
vectemp <- cat("The class of asked variable", column_name, "is", class(dfn[[column_name]]))
}
```
```{r}
summarize_column(df_hd, "age")

```
```{r}
summarize_column(df_hd, "diagnosed")

```
```{r error=TRUE}
summarize_column(df_hd, "random")

```
Test for the chr_20 database



